A generalization of the Bradley-Terry model for draws in chess

### 2019-05-23 inst/collusion_chess_draws.Rnw
various changes suggested by Jeff Nijsse.  Thanks Jeff\!

In inference problems where the dataset comprises Bernoulli outcomes
of paired comparisons, the Bradley-Terry model offers a simple and
easily interpreted framework.  However, it does not deal easily with
chess because of the existence of draws, and the white player
advantage.  Here I present a new generalization of Bradley-Terry in
which a chess game is regarded as a three-way competition between the
two players and an entity that wins if the game is drawn.
Bradley-Terry is then further generalized to account for the white
player advantage by positing an entity whose strength is added to that
of the white player.  These techniques afford insight into players'
strengths, response to playing black or white, and risk-aversion as
manifested by probability of drawing.  The likelihood functions
arising are easily optimised numerically.  I analyse a number of
datasets of chess results, including the infamous World Chess
Championships 1963 in which Fischer accused three Soviet players of
collusion.  I conclude, on the basis of a dataset that includes only
the scorelines at the event itself, that the Candidates Tournament
(Cura\c{c}ao 1962) exhibits evidence of collusion ($p<10^{-5}$), in
agreement with previous work that uses additional information such as
historical results including number of moves.}

@@

Many situations involve pairwise comparisons that amount to a
Bernoulli trial.  The canonical application would be to two-way
competitive situations such as tennis in which one or other player is
identified as the winner.  The Bradley-Terry model~\citep{bradley1952}
considers~$n$ players and assigns non-negative numbers
$p_1,\ldots,p_n$ with unit sum to each player.  Then player $i$ beats
player $j$ with probability $\frac{p_i}{p_i+p_j}$.  This simple system
has been modified and extended many times, for example to account for
multi-way competitions such as {\em MasterChef
  Australia}~\citep{hankin2017}, and the system has been applied to
many non-competitive situations such as consumer
behaviour~\citep{hankin2010} and mating preference~\citep{west2008}.

Bradley-Terry is not directly applicable to Chess for two reasons:
firstly, the white player appears to be at a considerable advantage;
and secondly, many chess games end in a draw: white wins about
$37.5\%$ of all games, black wins $27.6\%$, and about $39.4\%$ are
drawn.  We consider the first-mover advantage first, then consider the
effect of drawn games.

@@

\paragraph{Draws} Chess games may be won by either player, or may be
drawn; the implication of a draw is that neither player is better than
the other.  Before 1867, tournament games that were drawn were
replayed until one or other player won---reducing a chess game to a
Bernoulli trial (with probability one).  This system became unworkable
due to the large number of draws, resulting in the modern practice of
awarding each player of a drawn game half a point, keeping chess a
zero-sum game.  Drawn games are generally considered to be undesirable
as they are usually poorer in spectacular or striking elements such as
sharp tactical combinations or sacrifice.  Excessive drawing dissuades
sponsors, leaves spectators feeling cheated, and gives commentators
less material to discuss.  Players themselves do not seem to value
draws in the wider context of the game: \citeauthor{fischer1969}, for
example, includes only 8 draws in his compilation of 60 memorable
games, surely not representative of tournament play in which about
39.4\% of games are drawn (binomial test, $p<10^{-5}$).

\cite{moul2009} conceptualise a player as being able to adopt a
strategy on a spectrum ranging from conservative to aggressive, thus
being able to exert some control over the probability of a draw and
suggest that risk-averse players may pursue less vigorous strategies,
trading expected payout for reduced variance.  Further,
\cite{langer2013} point out that rational players in an ongoing
competitive situation modify their strategies depending on whether
they are leading or lagging: the lagging player will adopt a higher
risk strategy (`nothing to lose'), while the leading player is
incentivised to protect the lead by lowering the risk of losing.  In a
chess context, this means the leading player will aim for a draw.

@@

Consider first the standard Bradley-Terry model for a finite number of
chess players.  First we ignore the effect of playing white.  We wish
to find numbers (``strengths") $p_i$ for $1\leqslant i\leqslant n$
with $p_i\geqslant 0, \sum p_i=1$ such that the probability of $i$
beating $j\neq i$ is $\frac{p_i}{p_i+p_j}$.  A suitable likelihood
function, on the assumption of independence of trials, would be

@@

where $W_{ij}$ is the number of times $i$ beats $j$ and $L_{ij}$ is
the number of times $i$ loses to $j$.  Here I present a new
generalization of Bradley-Terry in which a chess game is regarded as a
three-way competition between the two players and an entity that wins
if the game is drawn.  The system is further modified to account for
first-move advantage by positing an entity whose strength is added to
that of the white player.  This methodology affords insight into
players' strengths, response to playing black or white, and
risk-aversion as manifested by probability of drawing.  If we consider
a chess game to be a three-way trial, an appropriate likelihood
function for non-negative $p_1,\ldots,p_n,D$ with $D+\sum p_i=1$ would
be

@@

where $W_{ij}$ is the number of games won by $i$ when playing $j$,
$L_{ij}$ is games lost and $D_{ij}$ is games drawn.  However, this
approach ignores the extra strength conferred by the first-move
advantage.  We thus consider another entity whose strength is added to
the white player.  The strength of the white player is thus the sum of
the player's strength and another entity with strength $W$.  Recalling
that a game might be a draw, we obtain the probabilities as given in
Table~\ref{probabilities}.  The corresponding likelihood function is
given below in equation \ref{likelihood_WD}:

@@

which might be preferable on grounds of clarity.  We now apply this
and related likelihood models to real chess datasets using the Method
of Support~\citet{edwards1972}.  To test any hypothesis $H$ on
likelihood function $\mathcal{L}$ one first identifies the constraint
on hypothesis space corresponding to $H$.  One then finds the
unconstrained maximum likelihood $\mathcal{L}_\mathrm{free}$ and
compares it with the maximum likelihood $\mathcal{L}_\mathrm{cons}$
when the optimization is constrained by $H$.  The test statistic is
then the likelihood ratio
$\mathcal{L}_\mathrm{free}/\mathcal{L}_\mathrm{cons}$ or, equivalently
the support $\mathcal{S}=\log\mathcal{R}$.  Wilks's theorem states
that the asymptotic distribution of $2\mathcal{S}$ is $\chi^2_n$ where
$n$ is the number of degrees of freedom implemented by $H$.

The above likelihood functions are objective and reasonable
representations of chess results.  However, they differ markedly from
previous likelihood functions of draws in that they include reified
entities representing proclivity to draw $D$, and white or first-mover
advantage ($W$).

@@

\cite{davidson1970} uses a threshhold approach to interpret a draw as
performance of the players the same to within a threshhold.  Using a
logistic expression for the difference in players' performance, he
effectively uses, in current notation, $D=\nu\sqrt{p_ip_j}$.  However,
\cite{joe1990} uses likelihood methods to show that this model does
not fit data well.  Here we note that such a model automatically give
stronger players a higher draw potential, and also note that it is
difficult to account for first-player advantage in this framework.

Such an approach is
numerically challenging in the current framework as the unit-sum
constraint becomes nonlinear and likelihood maximization becomes
difficult.

@@

The Method of Support \citep{edwards1972} involves only numerical
optimization (not integration which is computationally expensive).
The analysis presented here uses the framework of the \pkg{hyper2}
package~\citep{hankin2017} as a convenient and unified interface.
Because gradients are available, convergence to the evaluate is
generally rapid and hypothesis-specific constraints on parameter space
are, in general, straightforward to implement.

@@

Grandmasters Anand, Karpov, Kasparov have long-standing rivalry and
have played many games amongst themselves.  A summary may be found
in Table~\ref{kka}.  Equation~\ref{likelihood_WD} becomes

\begin{table}
\centering
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays White,}\\ \small{wins}} & \multicolumn{3}{c@{}}{\small{Plays black, loses}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  18 &   6  \\
  Kr &  7 &  -  &  18  \\
  Ks & 15 &  30 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{draws}} & \multicolumn{3}{c@{}}{\small{Plays black, draws}}\\
 \cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  20 &  20  \\
  Kr & 29 &  -  &  72  \\
  Ks & 26 &  57 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{loses}} & \multicolumn{3}{c@{}}{\small{Plays black, wins}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &   5 &   11  \\
  Kr & 13 &  -  &    9  \\
  Ks &  2 &   7 &  -   \\
\end{tabular}
\caption{Results of 365 chess games \label{kka} between Grandmaster
  Anand, Grandmaster Karpov, and Grandmaster Kasparov.  Thus, for
  example, Anand played white against Karpov a total of $18+20+13=51$
  times and won 18, drew 20, lost 13}
\end{table}

@@

\begin{figure}[t]
\begin{center}
<<ddc,echo=FALSE,fig=TRUE>>=
dotchart(maxp(karpov_kasparov_anand),pch=16,xlim=c(0,0.55))
@
\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
\label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_WD}}
  \end{center}
\end{figure}

This likelihood function is easily maximized numerically and this is
shown visually in figure~\ref{maxp_karpov_kasparov_anand}, which
suggests a number of interesting and testable statistical hypotheses.
Firstly, we observe that $D=0$ cannot possibly be the case: the
likelihood for this hypothesis would be identically zero, owing to the
nonzero term in the numerator of equation~\ref{kka_likelihood}.  From
the figure, we see that the maximum likelihood estimate for draw
strength $\hat{D}\simeq 0.53$: compare the na\"ive estimate of
$\frac{224}{365}\simeq 0.61$ given by observing 224 draws among 265
games.  The MLE is lower because the maximization process
automatically adjusts for the different players' strengths.  Secondly,
we may make inferences about $W$, the extra strength conferred by
playing white: it is conceivable that $W=0$ as this term does not
appear on its own in the numerator of any term.  We may test
$H_0\colon W=0$ using the Method of Support by maximizing the
likelihood subject the constraint $W=0$ and comparing with the
likelihood when the optimization is free.

@@

In figure~\ref{kka_3draws}, we see considerable variability in
proclivity to draw and again this may be tested with the Method of
Support, which gives a $p$-value of about 0.039, borderline
significant: an interesting result, given that in this case no draw
monster appears alone anywhere in the likelihood function.

@@

The Method of Support allows one to estimate proclivity to draw while
accounting for players' strengths and first-move advantage.  We now
consider one famous incident, following the World Chess Championships
1963, in which Fischer famously and publicly claimed that three of the
Soviet players colluded.  The accusation was that Keres, Petrosian,
and Geller agreed ahead of time to draw when playing one another, in
order to conserve energy for subsequent matches.

@@

Here I consider only the game results, reproduced in
table~\ref{candidates}.  We see that all twelve of the games between
the three accused players ended in a draw, a significant result given
51 winning games and 54 draws (Fisher exact test, one-sided,
$p=0.0079$).  To quantify such Soviet collusion, accounting for the
player skills, we introduce a new draw entity: a collusive draw
monster $C$, who wins when two colluding players draw, and is inactive
whenever a game has at least one non-colluding player.  Then
$H_N\colon C=D$ corresponds to no collusion and the likelihood
function would have

@@

\noindent on the numerator (where $C$ is the strength of a collusive
draw), and a total of 30 terms such as
$\left(P+\mathit{Ke}+W+C\right)^{4}$ [indicating that Petrosian played
  Keres 4 times] on the denominator.  The likelihood ratio for the
hypothesis that collusive draws have the same strength as regular
draws is about $e^{11.8}\simeq 10^5$: we conclude that the high number
of draws among the three accused players was unlikely to be due to
chance.

@@

The maximum likelihood estimate is $D\simeq 0.043, D_s=0.354$; a
constrained maximization gives a likelihood ratio of about 5096, or a
$p$-value of about $3.6\times 10^{-5}$.  Indeed we could go further
and plot a profile likelihood curve for the log-odds ratio
$\ln\left(D_s/D\right)$ which, if Dirichlet, would be accurately
parabolic over a wide range~\cite[page 343]{ohagan2004}.  This is
illustrated in Figure~\ref{proflike}, that shows a credible interval
of about 1.8-2.4.

@@

In this context we note, following \cite{langer2013}, that a player
may well adopt a riskier strategy after a sequence of defeats,
adopting a ``nothing-to-lose'' mentality.  Such a phenomenon would be
detectable, at least in principle, by generalizing the personalised
draw likelihood function \ref{KKA_personalised_drawmonster} and
seeking evidence for the draw strength changing as a tournament
progresses: it would become stronger for a player on a winning streak,
and weaker for a losing player.

>>>>

In inference problems where the dataset comprises Bernoulli outcomes
of paired comparisons, the Bradley-Terry model offers a simple and
easily interpreted framework.  However, it does not deal easily with
chess because of the existence of draws, and the white player
advantage.  Here I present a new generalization of Bradley-Terry in
which a chess game is regarded as a three-way competition between the
two players and an entity that wins if the game is drawn.
Bradley-Terry is then further generalized to account for the white
player advantage by positing a second entity whose strength is added
to that of the white player.  These techniques afford insight into
players' strengths, response to playing black or white, and
risk-aversion as manifested by probability of drawing.  The likelihood
functions arising are easily optimised numerically.  I analyse a
number of datasets of chess results, including the infamous 1963 World
Chess Championships, in which Fischer accused three Soviet players of
collusion.  I conclude, on the basis of a dataset that includes only
the scorelines at the event itself, that the Candidates Tournament
(Cura\c{c}ao 1962) exhibits evidence of collusion ($p<10^{-5}$), in
agreement with previous work that uses additional information such as
historical results including number of moves.}

@@

Many situations involve pairwise comparisons that amount to a
Bernoulli trial.  The canonical application would be two-way
competitive situations such as tennis in which one player is
identified as the winner.  The Bradley-Terry model~\citep{bradley1952}
considers~$n$ players and assigns non-negative numbers
$p_1,\ldots,p_n$ with unit sum to each player.  Then player $i$ beats
player $j$ with probability $\frac{p_i}{p_i+p_j}$.  This simple system
has been modified and extended many times, for example to account for
multi-way competitions such as {\em MasterChef
  Australia}~\citep{hankin2017}, and non-competitive situations such
as consumer behaviour~\citep{hankin2010} and mating
preference~\citep{west2008}.

Bradley-Terry is not directly applicable to chess for two reasons:
firstly, the white player appears to be at a considerable advantage
winning about 37.5\% of games (black wins 27.6\%); and secondly, many
chess games, about 39.4\%, are drawn (figures from Chessbase).

@@

\paragraph{Draws} Chess games may be won by either player, or may be
drawn; the implication of a draw is that neither player is better than
the other.  Before 1867, tournament games that were drawn were
replayed until one or other player won---reducing a chess game to a
Bernoulli trial (with probability one).  This system became unworkable
due to the large number of draws, resulting in the modern practice of
awarding each player of a drawn game half a point, keeping chess a
zero-sum game.  Drawn games are generally considered to be undesirable
as they are usually poorer in spectacular or striking elements such as
sharp tactical combinations or sacrifice.  Excessive drawing dissuades
sponsors, leaves spectators feeling cheated, and gives commentators
less material to discuss.  Players themselves do not seem to value
draws in the wider context of the game: \citet{fischer1969}, for
example, includes only 8 draws in his compilation of 60 memorable
games, surely not representative of tournament play in which about
39.4\% of games are drawn (binomial test, $p<10^{-5}$).

\cite{moul2009} conceptualise a player as being able to adopt a
strategy on a spectrum ranging from conservative to aggressive, thus
being able to exert some control over the probability of a draw and
suggest that risk-averse players may pursue less vigorous strategies,
trading expected payout for reduced variance.  Further,
\cite{langer2013} point out that rational players in an ongoing
competitive situation modify their strategies depending on whether
they are leading or lagging: the lagging player will adopt a higher
risk strategy (`nothing to lose'), while the leading player is
incentivised to protect the lead by lowering the risk of losing.  In a
chess context, this means the leading player will be more likely to
aim for a draw.

@@

Consider first the standard Bradley-Terry model for a finite number of
chess players.  Ignoring the effect of playing white, we wish to find
numbers (``strengths") $p_i$ for $1\leqslant i\leqslant n$ with
$p_i\geqslant 0, \sum p_i=1$ such that the probability of $i$ beating
$j\neq i$ is $\frac{p_i}{p_i+p_j}$.  A suitable likelihood function,
on the assumption of independence of trials, would be

@@

where $W_{ij}$ is the number of times $i$ beats $j$ and $L_{ij}$ is
the number of times $i$ loses to $j$.  Here I present a new
generalization of Bradley-Terry in which a chess game is regarded as a
three-way competition between the two players and a third entity (a
``draw monster'') that wins if the game is drawn.  The system is
further modified to account for first-move advantage by positing an
entity whose strength is added to that of the white player.  This
methodology affords insight into players' strengths, response to
playing black or white, and risk-aversion as manifested by probability
of drawing.  If we consider a chess game to be a three-way trial, an
appropriate likelihood function would be for non-negative
$p_1,\ldots,p_n,D$ ($D$ being the draw monster's strength), with
$D+\sum p_i=1$ and

@@

where $W_{ij}$ is the number of games won by $i$ when playing $j$,
$L_{ij}$ is games lost and $D_{ij}$ is games drawn.  However, this
approach ignores the extra strength conferred by the first-move
advantage.  We thus consider another entity whose strength $W$ is
added to the white player.  The strength of the white player is thus
the sum of the player's strength and $W$.  Recalling that a game might
be a draw, we obtain the probabilities as given in
table~\ref{probabilities}.  The corresponding likelihood function is
given below in equation \ref{likelihood_WD}:

@@

which might be preferable on grounds of clarity.  We now apply this
and related likelihood models to real chess datasets using the Method
of Support~\citep{edwards1972}.  To test any hypothesis $H$ on
likelihood function $\mathcal{L}$ one first identifies the constraint
on hypothesis space corresponding to $H$.  One then finds the
unconstrained maximum likelihood $\mathcal{L}_\mathrm{free}$ and
compares it with the maximum likelihood $\mathcal{L}_\mathrm{cons}$
when the optimization is constrained by $H$.  The test statistic is
then the likelihood ratio
$\mathcal{L}_\mathrm{free}/\mathcal{L}_\mathrm{cons}$ or, equivalently
the support $\mathcal{S}=\log\mathcal{R}$.  Wilks's theorem states
that the asymptotic distribution of $2\mathcal{S}$ is $\chi^2_n$ where
$n$ is the number of degrees of freedom implemented by $H$.

The above likelihood functions are objective and reasonable
representations of chess game outcomes.  However, they differ markedly
from previous likelihood functions of draws in that they include
reified entities representing proclivity to draw $D$, and white or
first-mover advantage ($W$).

@@

\cite{davidson1970} uses a threshhold approach to interpret draws.
From a chess draw, he infers that the performance of the players is
identical, within a threshhold.  Using a logistic expression for the
difference in players' performance he uses, in current notation,
$D=\nu\sqrt{p_ip_j}$.  However, \cite{joe1990} uses likelihood methods
to show that this model does not fit data well.  Here we note that
such a model automatically give stronger players a higher draw
potential, and also note that it is difficult to account for
first-player advantage in this framework.  Such an approach is
numerically challenging in the current framework as the unit-sum
constraint becomes nonlinear, making likelihood maximization
difficult.  \cite{moul2009} consider chess draws and use sophisticated
regression techniques to analyse games in terms of not only the
result, but also the ply-count, data not used in the present analysis.

@@

The Method of Support \citep{edwards1972} involves only numerical
optimization (not integration which is computationally expensive).
The analysis presented here uses the framework of the
R~\citep{rcore2019} \pkg{hyper2} package~\citep{hankin2017} as a
convenient and unified interface.  Because gradients are available,
convergence to the evaluate is generally rapid and hypothesis-specific
constraints on parameter space are, in general, linear and
straightforward to implement.

@@

Grandmasters Anand, Karpov, and Kasparov have long-standing rivalry
and have played many games amongst themselves.  A summary may be found
in table~\ref{kka}.  Equation~\ref{likelihood_WD} becomes

\begin{table}
\centering
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays White,}\\ \small{wins}} & \multicolumn{3}{c@{}}{\small{Plays black, loses}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  18 &   6  \\
  Kr &  7 &  -  &  18  \\
  Ks & 15 &  30 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{draws}} & \multicolumn{3}{c@{}}{\small{Plays black, draws}}\\
 \cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  20 &  20  \\
  Kr & 29 &  -  &  72  \\
  Ks & 26 &  57 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{loses}} & \multicolumn{3}{c@{}}{\small{Plays black, wins}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &   5 &   11  \\
  Kr & 13 &  -  &    9  \\
  Ks &  2 &   7 &  -   \\
\end{tabular}
\caption{Results of 365 chess games \label{kka} between Grandmaster
  Anand, Grandmaster Karpov, and Grandmaster Kasparov.  For example,
  Anand played white against Karpov a total of $18+20+13=51$ times and
  won 18, drew 20, lost 13}
\end{table}

@@

\begin{figure}[t]
\begin{center}
<<ddc,echo=FALSE,fig=TRUE>>=
dotchart(maxp(karpov_kasparov_anand),pch=16,xlim=c(0,0.55))
@
\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in table~\ref{kka} 
\label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_WD}}
  \end{center}
\end{figure}

This likelihood function is easily maximized numerically and the
evaluate is shown visually in figure~\ref{maxp_karpov_kasparov_anand},
which suggests a number of interesting and testable statistical
hypotheses.  Firstly, we observe that $D=0$ cannot possibly be the
case: the likelihood for this hypothesis would be identically zero,
owing to the nonzero term in the numerator of
equation~\ref{kka_likelihood}.  From the figure, we see that the
maximum likelihood estimate for draw strength $\hat{D}\simeq 0.53$:
compare the na\"ive estimate of $\frac{224}{365}\simeq 0.61$ given by
observing 224 draws among 265 games.  The na\"{i}ve estimate is lower
because the maximization process automatically adjusts for the
different players' strengths.  Secondly, we may make inferences about
$W$, the extra strength conferred by playing white: it is conceivable
that $W=0$ as this term does not appear on its own in the numerator of
any term.  We may test $H_0\colon W=0$ using the Method of Support by
maximizing the likelihood subject to the constraint $W=0$ and
comparing with the likelihood when the optimization is free.

@@

In figure~\ref{kka_3draws}, we see considerable variability in
proclivity to draw.  Again this may be tested with the Method of
Support, which gives a $p$-value of about 0.039, borderline
significant: an interesting result, given that in this case no draw
monster appears alone anywhere in the likelihood function.

@@

The Method of Support allows one to estimate proclivity to draw while
accounting for players' strengths and first-move advantage.  We now
consider one infamous incident, following the World Chess
Championships 1963, in which Fischer famously and publicly claimed
that three of the Soviet players colluded.  The accusation was that
Keres, Petrosian, and Geller agreed ahead of time to draw when playing
one another, in order to conserve energy for subsequent matches.

@@

Here I consider only the game results, reproduced in
table~\ref{candidates}.  We see that all twelve of the games between
the three accused players ended in a draw, a significant result given
51 winning games and 54 draws (Fisher's exact test, one-sided,
$p=0.0079$).  To quantify such Soviet collusion, accounting for the
player skills, we introduce a new draw entity: a collusive draw
monster $C$, who wins when two colluding players draw, and is inactive
whenever a game has at least one non-colluding player.  Then
$H_N\colon C=D$ corresponds to no collusion and the likelihood
function would have

@@

\noindent for the numerator (where $C$ is the strength of a collusive
draw), and a total of 30 terms such as
$\left(P+\mathit{Ke}+W+C\right)^{4}$ [indicating that Petrosian played
  Keres 4 times] in the denominator.  The likelihood ratio for the
hypothesis that collusive draws have the same strength as regular
draws is about $e^{11.8}\simeq 10^5$: we conclude that the high number
of draws among the three accused players was unlikely to be due to
chance.

@@

The maximum likelihood estimate is $D\simeq 0.043, D_s=0.354$; a
constrained maximization gives a likelihood ratio of about 5096, or a
$p$-value of about $3.6\times 10^{-5}$.  Indeed we could go further
and plot a profile likelihood curve for the log-odds ratio
$\log\left(D_s/D\right)$ which, if Dirichlet, would be accurately
parabolic over a wide range~\cite[page 343]{ohagan2004}.  This is
illustrated in figure~\ref{proflike}, that shows a credible interval
of about 1.8-2.4.

@@

In this context we note, following the observation of~\citet{moul2009}
that a player may well adopt a riskier strategy after a sequence of
defeats, adopting a ``nothing-to-lose'' mentality.  Such a phenomenon
would be detectable, at least in principle, by generalizing the
personalised draw likelihood function
\ref{KKA_personalised_drawmonster} and seeking evidence for the draw
strength changing as a tournament progresses: it would become stronger
for a player on a winning streak, and weaker for a losing player.
